---
title: "Real Estate Data Analysis"

subtitle: "RAM0580 Andmeanalüüs - Serminar 1"

author: "<br>Authors: Erik Schults, Egert Blok <br>"

date: "Date: 2024.02.12 <br>"


output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float: 
      collapsed: false
    number_sections: true
    theme: default
    code_folding: show
---

Housing dataset from https://www.kaggle.com/datasets/vikramamin/housing-linear-regression

# Data import

```{r}
housing <- read.csv("housing.csv", dec = ".", stringsAsFactors = FALSE)
```


```{r}
library(skimr)
skim(housing)
```

```{r, results='asis'}
library(summarytools)
dfSummary(housing,plain.ascii = FALSE, style = "grid",tmp.img.dir  = "/tmp",graph.magnif = 0.85)
```

## Missing values

```{r}
sum(is.na(housing))
sum(!complete.cases(housing))
apply(is.na(housing), 2, sum)
```

## Conclusion

No empty rows found.
The address can be discarded as it does not provide additional information.
No factors need to be formed as all values are numeric.

The address is removed from the set because all values are distinct strings.

```{r}
library(dplyr)
housing <-housing %>%
    select(c(
    "Avg..Area.Income",
    "Avg..Area.House.Age",
    "Avg..Area.Number.of.Rooms",
    "Avg..Area.Number.of.Bedrooms",
    "Area.Population",
    "Price"
  ))
```

# Descriptive data analysis

The descriptive data analysis consists of presenting:

* Diagrams of the distribution of characteristics to show how the values of certain
  (non)numeric characteristic is spread or distributed across the dataset. The
  diagrams can include histograms, density plots, and box plots which help understanding
  the tendency, dispersion, and shape of the data distribution.
    * Tendency: A central or typical value of a dataset (e.g avg, mean, median, mode).
    * Dispersion(variability): Shows how spread out the values in a dataset are compared to
      the central point.
      This includes the range, interquartile range (IQR, range of the middle 50,75%),
      variance, and standard deviation.
        * Range: The difference between the highest and lowest points. It gives
          a quick sense about the spread, but is heavily influence by the outliers.
        * Variance: Average of the squared difference from the mean.
          Provides good insights into the variability within a dataset, by considering
          how each datapoint differs form the mean.
        * Standard deviation: Sqrt of variance, indicating that the datapoints
          are about x (std) units from the mean. It provides insights into the dispersion of the data.
* Frequency tables to identify the most (un)common values and to understand
  the distribution of categorical data.

## Price

The target property _Y=Price_ is a continuous variable.

Let's visualize the distribution using a histogram.


```{r}
Price_hist <- hist(
  housing$Price, 
  col = "skyblue",
  main = "Price frequency",
  xlab="Price",
  ylab = "Frequency"
)

```

The relative distribution:

```{r}
par(mfrow=c(1,2))

Price_hist$counts <- Price_hist$counts / sum(Price_hist$counts) * 100

plot(
  Price_hist,
  col = "skyblue",
  main = "Price % distribution",
  xlab="Price",
  ylab = "Frequency %",
  cex.main=0.8
)

Price_box <- boxplot(
  housing$Price, 
  col = "skyblue",
  horizontal = 1,
  main = "Price box plot",
  xlab = "Price",
  cex.main = 0.8,
  range = 2
)
```

_Price_ has outliers on both sides are:

```{r}
sort(Price_box$out)
```
The 2 outliers on the left side have the values of 16e3 and 31e3.
The 1 outlier on the right has a value of 2.5e6;
Details of the 3 outliers:

```{r}
Price_out <- which(housing$Price %in% Price_box$out)

library(DT)
datatable(
  housing[Price_out, ],
  options=list(scrollX=1,pageLength=3,searching = FALSE,scroller = TRUE,scrollY=200)
)
```

Frequency table of _Price_ (first the data has to be _cut()_ into intervals based on the histogram).

```{r}
intervals <- cut(housing$Price, Price_hist$breaks, include.lowest = TRUE)
col_frequency <- table(intervals)
col_frequency_rel <- prop.table(col_frequency) * 100
Price_ftable <- cbind(rownames(col_frequency), col_frequency, round(col_frequency_rel, 2), round(cumsum(col_frequency_rel), 2))
Price_ftable <- rbind(Price_ftable, c("Total", sum(col_frequency), sum(col_frequency_rel), ""))
Price_ftable <- data.frame(Price_ftable, row.names = NULL)
names(Price_ftable) <- c("Price", "Frequency", "Frequency %", "Cumulative f. %")

library(knitr)
library(kableExtra)

kable_styling(
  kable(
    Price_ftable,
    align = c('l','r','r', 'r')
  ),
  bootstrap_options = c("striped","hover"),
  full_width = 0,
  position = "left"
)
```

Numerical characteristics of the variable _Price_:

```{r}
library(FSA)

Price_summary <- Summarize(housing$Price, digits = 2)
kable_styling(kable(t(Price_summary)))
```

_Price_ is quite symmetric as indicated by its mean (1232073) marginally exceeding the median (1232669).
A majority, 42%, of the _Price_ observations are clustered within a 1e6-1.5e6 range. The dataset of this variable
ranges from a minimum of 15e3 to a maxiumum of 2.4e6.



## Avg..Area.Income

The target property _Y=Avg..Area.Income_ is a continuous variable.

Let's visualize the distribution using a histogram.


```{r}
Avg..Area.Income_hist <- hist(
  housing$Avg..Area.Income, 
  col = "skyblue",
  main = "Avg income frequency",
  xlab="Avg income",
  ylab = "Frequency",
  breaks = 12
)

```

The relative distribution:

```{r}
par(mfrow=c(1,2))

Avg..Area.Income_hist$counts <- Avg..Area.Income_hist$counts / sum(Avg..Area.Income_hist$counts) * 100

plot(
  Avg..Area.Income_hist,
  col = "skyblue",
  main = "Avg income % distribution",
  xlab="Avg income",
  ylab = "Frequency %",
  cex.main=0.8
)

Avg..Area.Income_box <- boxplot(
  housing$Avg..Area.Income, 
  col = "skyblue",
  horizontal = 1,
  main = "Avg income box plot",
  xlab = "Avg income",
  cex.main = 0.8,
  range = 1.9
)
```

_Avg..Area.Income_hist_ has an asymmetric left side. The outliers are:

```{r}
sort(Avg..Area.Income_box$out)
```
The outlier on the left side has a value of ~17769 and the values of the 2 outliers on right
are 104e3 and 107e3.
Details of the 3 outliers:

```{r}
Avg..Area.Income_out <- which(housing$Avg..Area.Income %in% Avg..Area.Income_box$out)

library(DT)
datatable(
  housing[Avg..Area.Income_out, ],
  options=list(scrollX=1,pageLength=3,searching = FALSE,scroller = TRUE,scrollY=200)
)
```


Frequency table of _Avg..Area.Income_.

```{r}
intervals <- cut(housing$Avg..Area.Income, Avg..Area.Income_hist$breaks, include.lowest = TRUE)
col_frequency <- table(intervals)
col_frequency_rel <- prop.table(col_frequency) * 100
Avg..Area.Income_ftable <- cbind(rownames(col_frequency), col_frequency, round(col_frequency_rel, 2), round(cumsum(col_frequency_rel), 2))
Avg..Area.Income_ftable <- rbind(Avg..Area.Income_ftable, c("Total", sum(col_frequency), sum(col_frequency_rel), ""))
Avg..Area.Income_ftable <- data.frame(Avg..Area.Income_ftable, row.names = NULL)
names(Avg..Area.Income_ftable) <- c("Avg..Area.Income", "Frequency", "Frequency %", "Cumulative f. %")

library(knitr)
library(kableExtra)

kable_styling(
  kable(
    Avg..Area.Income_ftable,
    align = c('l','r','r', 'r')
  ),
  bootstrap_options = c("striped","hover"),
  full_width = 0,
  position = "left"
)
```

Numerical characteristics of the variable _Avg..Area.Income_:

```{r}
library(FSA)

Avg..Area.Income_summary <- Summarize(housing$Avg..Area.Income, digits = 2)
kable_styling(kable(t(Avg..Area.Income_summary)))
```

_Avg..Area.Income_ is quite symmetric, with a slight left skew, as indicated by its mean (68583) marginally deceeding the median (68804).
A majority, 65%, of the _Avg..Area.Income_ observations are clustered within a 60k to 80k range.
The dataset of this variable ranges from a minimum of 17e3 to a maxiumum of 107e3.


## Avg..Area.House.Age

The target property _Y=Avg..Area.House.Age_ is a continuous variable.

Let's visualize the distribution using a histogram.


```{r}
Avg..Area.House.Age_hist <- hist(
  housing$Avg..Area.House.Age, 
  col = "skyblue",
  main = "Avg house age frequency",
  xlab="Avg house age",
  ylab = "Frequency"
)

```

The relative distribution:

```{r}
par(mfrow=c(1,2))

Avg..Area.House.Age_hist$counts <- Avg..Area.House.Age_hist$counts / sum(Avg..Area.House.Age_hist$counts) * 100

plot(
  Avg..Area.House.Age_hist,
  col = "skyblue",
  main = "Avg house age % distribution",
  xlab="Avg house age",
  ylab = "Frequency %",
  cex.main=0.8
)

Avg..Area.House.Age_box <- boxplot(
  housing$Avg..Area.House.Age, 
  col = "skyblue",
  horizontal = 1,
  main = "Avg house age box plot",
  xlab = "Avg house age",
  cex.main = 0.8,
  range = 1.6
)
```

_Avg..Area.House.Age_ has a slight asymmetric right side. The outliers are:

```{r}
sort(Avg..Area.House.Age_box$out)
```

The outliers are in the ranges of ~2.64 to ~3.14 (in total 10) and ~8.92 to ~9.51 (in total 7).
Details of the 17 outliers:

```{r}
Avg..Area.House.Age_out <- which(housing$Avg..Area.House.Age %in% Avg..Area.House.Age_box$out)

library(DT)
datatable(
  housing[Avg..Area.House.Age_out, ],
  options=list(scrollX=1,pageLength=7,searching = FALSE,scroller = TRUE,scrollY=200)
)
```


## Avg..Area.Number.of.Rooms

The target property _Y=Avg..Area.Number.of.Rooms_ is a continuous variable.

Let's visualize the distribution using a histogram.


```{r}
Avg..Area.Number.of.Rooms_hist <- hist(
  housing$Avg..Area.Number.of.Rooms, 
  col = "skyblue",
  main = "Avg rooms frequency",
  xlab="Avg rooms",
  ylab = "Frequency"
)

```

The relative distribution:

```{r}
par(mfrow=c(1,2))

Avg..Area.Number.of.Rooms_hist$counts <- Avg..Area.Number.of.Rooms_hist$counts / sum(Avg..Area.Number.of.Rooms_hist$counts) * 100

plot(
  Avg..Area.Number.of.Rooms_hist,
  col = "skyblue",
  main = "Avg rooms % distribution",
  xlab="Avg rooms",
  ylab = "Frequency %",
  cex.main=0.8
)

Avg..Area.Number.of.Rooms_box <- boxplot(
  housing$Avg..Area.Number.of.Rooms, 
  col = "skyblue",
  horizontal = 1,
  main = "Avg rooms box plot",
  xlab = "Avg rooms",
  cex.main = 0.8,
  range = 1.65
)
```

_Avg..Area.Number.of.Rooms_ has a slight asymmetric left side. The outliers are:

```{r}
sort(Avg..Area.Number.of.Rooms_box$out)
```

The outliers are in the ranges of ~3.23 to ~4.03 and ~9.93 to ~10.6.
Details of the outliers:

```{r}
Avg..Area.Number.of.Rooms_out <- which(housing$Avg..Area.Number.of.Rooms %in% Avg..Area.Number.of.Rooms_box$out)

library(DT)
datatable(
  housing[Avg..Area.House.Age_out, ],
  options=list(scrollX=1,pageLength=7,searching = FALSE,scroller = TRUE,scrollY=200)
)
```


## Avg..Area.Number.of.Bedrooms

The target property _Y=Avg..Area.Number.of.Bedrooms_ is a continuous variable.

Let's visualize the distribution using a histogram.


```{r}
Avg..Area.Number.of.Bedrooms_hist <- hist(
  housing$Avg..Area.Number.of.Bedrooms, 
  col = "skyblue",
  main = "Avg bedrooms frequency",
  xlab="Avg bedrooms",
  ylab = "Frequency",
  breaks = c(2, 3, 4, 5, 6, 7)
)

```

The relative distribution:

```{r}
par(mfrow=c(1,2))

Avg..Area.Number.of.Bedrooms_hist$counts <- Avg..Area.Number.of.Bedrooms_hist$counts / sum(Avg..Area.Number.of.Bedrooms_hist$counts) * 100

plot(
  Avg..Area.Number.of.Bedrooms_hist,
  col = "skyblue",
  main = "Avg bedrooms % distribution",
  xlab="Avg bedrooms",
  ylab = "Frequency %",
  cex.main=0.8
)

Avg..Area.Number.of.Bedrooms_box <- boxplot(
  housing$Avg..Area.Number.of.Bedrooms, 
  col = "skyblue",
  horizontal = 1,
  main = "Avg bedrooms box plot",
  xlab = "Avg bedrooms",
  cex.main = 0.8,
  range = 1.5
)
```

_Avg..Area.Number.of.Bedrooms_ has a slight asymmetric right side. No outliers were detected:

```{r}
sort(Avg..Area.Number.of.Bedrooms_box$out)
```



## Area.Population

The target property _Y=Area.Population_ is a continuous variable.

Let's visualize the distribution using a histogram.

```{r}
Area.Population_hist <- hist(
  housing$Area.Population, 
  col = "skyblue",
  main = "Area population frequency",
  xlab="Population",
  ylab = "Frequency"
)

```

The relative distribution:

```{r}
par(mfrow=c(1,2))

Area.Population_hist$counts <- Area.Population_hist$counts / sum(Area.Population_hist$counts) * 100

plot(
  Area.Population_hist,
  col = "skyblue",
  main = "Area population % distribution",
  xlab="Population",
  ylab = "Frequency %",
  cex.main=0.8
)

Area.Population_box <- boxplot(
  housing$Area.Population, 
  col = "skyblue",
  horizontal = 1,
  main = "Area population box plot",
  xlab = "Population rooms",
  cex.main = 0.8,
  range = 1.5
)
```

_Area.Population_ has a slight asymmetric left side. The outliers are:

```{r}
sort(Area.Population_box$out)
```

The outliers are in the ranges of ~170 to ~9200 and ~63000 to ~7000.
Details of the outliers:

```{r}
Area.Population_out <- which(housing$Area.Population %in% Area.Population_box$out)

library(DT)
datatable(
  housing[Area.Population_out, ],
  options=list(scrollX=1,pageLength=7,searching = FALSE,scroller = TRUE,scrollY=200)
)
```
## All outliers

In total 93 outliers were found.

```{r}
all_out <- c(Price_out, Avg..Area.Income_out, Avg..Area.Number.of.Rooms_out, Avg..Area.House.Age_out, Area.Population_out)

library(DT)
datatable(
  housing[all_out, ],
  options=list(scrollX=1,pageLength=7,searching = FALSE,scroller = TRUE,scrollY=200)
)
```
# Analysis

```{r}
# Total of 90 outliers were removed from the _housing_ dataset.
# housing <- housing[-all_out, ]
```

The target variable for analysis is _Price_.
Visualization of all the numeric variable distributions with a _plot()_.

```{r}
plot(Filter(is.numeric, housing), pch=20, col='blue', cex=.5)
```

Looks like The price grows linearly to all variables except _Avg..Area.Number.Of.Bedrooms_, but everything is quite scattered.
The independant variables data is scattered, but is still linearly growing. 
The most promising/clean linear incline to _Price_ happens with _Avg..Area.Income_.

## Avg..Area.Income

Looking at the relationship between _Avg..Area.Income_ and _Price_ using a _ggplot_.

```{r}
library(ggplot2)
housing %>%
  ggplot(aes(x = Price, y = Avg..Area.Income)) +
  geom_point() + 
  xlab("Price ($)") + 
  ylab("Avg Area Income ($/year)")
```

We can see that the _Price_ increases when  _Avg..Area.Income_ is increasing.
Let's add a color to _Area.Population_.

```{r}
library(ggplot2)
housing %>%
  ggplot(aes(x = Price, y = Avg..Area.Income, color = Area.Population)) +
  geom_point() + 
  xlab("Price ($)") + 
  ylab("Avg Area Income ($/year)")
housing %>%
  ggplot(aes(y = Price, x = Avg..Area.Income, color = Area.Population)) +
  geom_point() + 
  ylab("Price ($)") + 
  xlab("Avg Area Income ($/year)")
```
We can see that the price increases when  _Avg..Area.Income_ is increasing.
In lower population areas the price is lower.

Plots of filtered data:

```{r}
library(dplyr)

# TODO: find all permutations of independent vars to find homoscedastic relations 
# intervals for every independant variable
# x loop through each independant variable and each interval
# y loop through anothe independant variable

plot(Filter(is.numeric, housing) %>%
  filter(between(Avg..Area.Number.of.Rooms, 3, 5)) %>%
  select(c(
    "Price",
    "Avg..Area.Income",
    "Avg..Area.House.Age",
    "Area.Population"
  )), pch=20, col='blue', cex=.5, main="3-5 rooms")

plot(Filter(is.numeric, housing) %>%
  filter(between(Avg..Area.Number.of.Rooms, 5, 6)) %>%
  select(c(
    "Price",
    "Avg..Area.Income",
    "Avg..Area.House.Age",
    "Area.Population"
  )), pch=20, col='blue', cex=.5, main="5-6 rooms")

plot(Filter(is.numeric, housing) %>%
  filter(between(Avg..Area.Number.of.Rooms, 6, 7)) %>%
  select(c(
    "Price",
    "Avg..Area.Income",
    "Avg..Area.House.Age",
    "Area.Population"
  )), pch=20, col='blue', cex=.5, main="6-7 rooms")

plot(Filter(is.numeric, housing) %>%
  filter(between(Avg..Area.Number.of.Rooms, 7, 8)) %>%
  select(c(
    "Price",
    "Avg..Area.Income",
    "Avg..Area.House.Age",
    "Area.Population"
  )), pch=20, col='blue', cex=.5, main="7-8 rooms")


plot(Filter(is.numeric, housing) %>%
  filter(round(Avg..Area.Number.of.Rooms) == 7) %>%
  select(c(
    "Price",
    "Avg..Area.Income",
    "Avg..Area.House.Age",
    "Area.Population"
  )), pch=20, col='blue', cex=.5, main="7 rooms")


plot(Filter(is.numeric, housing) %>%
  filter(between(Avg..Area.Number.of.Rooms, 9, 11)) %>%
  select(c(
    "Price",
    "Avg..Area.Income",
    "Avg..Area.House.Age",
    "Area.Population"
  )), pch=20, col='blue', cex=.5, main="9-11 rooms")


```





See on histograms, how the data is distributed.


```{r}
price_hist <- hist(
  housing$Price, 
  col = "skyblue",
  main = "Price frequency",
  xlab="$",
  ylab = "Count"
)

par(mfrow=c(1,2))
price_hist$counts <- price_hist$counts / sum(price_hist$counts) * 100

price_hist$counts

plot(
  price_hist,
  col = "skyblue",
  main = "Price % distribution",
  xlab="$",
  ylab = "Frequency %",
  cex.main=0.8
)
```

```{r}

price_box <- boxplot(
  housing$Price, 
  col = "skyblue",
  horizontal = 1,
  main = "Price box plot",
  xlab = "$",
  range = 1.5,
  cex.main = 0.8
)

price_box_out <- which(housing$Price %in% price_box$out)
price_box_out

library(DT)
datatable(
  housing[price_box_out, ],
  options=list(scrollX=1,pageLength=10,searching = FALSE,scroller = TRUE,scrollY=200)
)
```



## Avg. Area House Age

```{r}
avg_age_hist <- hist(
  housing$Avg..Area.House.Age, 
  col = "skyblue",
  main = "House age frequency",
  xlab="Years",
  ylab = "Count"
)

```

Also the relative distribution

```{r}
par(mfrow=c(1,2))
avg_age_hist$counts <- avg_age_hist$counts / sum(avg_age_hist$counts) * 100

avg_age_hist$counts

plot(
  avg_age_hist,
  col = "skyblue",
  main = "House age % distribution",
  xlab="Years",
  ylab = "Frequency %",
  cex.main=0.8
)

avg_rooms_box <- boxplot(
  housing$Avg..Area.House.Age, 
  col = "skyblue",
  horizontal = 1,
  main = "House age box plot",
  xlab = "Years",
  range = 2,
  cex.main = 0.8
)
boxplot
```

Outliers of the __Avg..Area.House.Age__.

```{r}
sort(avg_rooms_box$out)
avg_rooms_box_out <- which(housing$Avg..Area.House.Age %in% avg_rooms_box$out)
avg_rooms_box_out

library(DT)
datatable(
  housing[avg_rooms_box_out, ],
  options=list(scrollX=1,pageLenght=5,searching = FALSE,scroller = TRUE,scrollY=200)
)
```



# Avg. Area Income

Lets compare the _Price_ with the discrete variables using box plots. The number of rooms
and the age of the house are discrete if rounded.

```{r}
par(mfrow=c(1,2))
discrete_vars = c(
  "Avg..Area.House.Age",
  "Avg..Area.Number.of.Rooms",
  "Avg..Area.Number.of.Bedrooms",
  "Area.Population",
  "Avg..Area.Income"
)

for (i in 1:length(discrete_vars)) {
  boxplot(
    housing$Price~round(housing[,discrete_vars[i]]),
    col="skyblue",
    xlab ="",
    ylab ="",
    las=2,
    cex.axis=0.5,
    cex.lab=0.7, 
    main=paste("Price vs" ,discrete_vars[i]),
    range=2
  )
}
```

# Linear regression

Let's get the summary for all of our variables with Price being our dependent variable and the other variables being independent.

```{r}

model <- lm(Price ~ Avg..Area.Income + Avg..Area.House.Age + Avg..Area.Number.of.Rooms + Avg..Area.Number.of.Bedrooms + Area.Population, data = housing)

summary(model)
```
We can see that "Average Area Income" coefficient does indeed have the highest effect on price, followed by "Average Area House Age" and "Area Population".

Let's show graphs for all variables:

```{r}
plot(housing$Avg..Area.Income, housing$Price, main="House Price vs Income", xlab="Income", ylab="House Price")
model_HouseIncome <- lm(Price ~ Avg..Area.Income, data = housing)
abline(model_HouseIncome, col="red")
```

```{r}
plot(housing$Avg..Area.House.Age, housing$Price, main="House Price vs House Age", xlab="House Age", ylab="House Price")
model_HouseAge <- lm(Price ~ Avg..Area.House.Age, data = housing)
abline(model_HouseAge, col="red")
```

```{r}
plot(housing$Avg..Area.Number.of.Rooms, housing$Price, main="House Price vs Rooms", xlab="Rooms", ylab="House Price")
model_HouseRooms <- lm(Price ~ Avg..Area.Number.of.Rooms, data = housing)
abline(model_HouseRooms, col="red")
```

```{r}
plot(housing$Avg..Area.Number.of.Bedrooms, housing$Price, main="House Price vs Bedrooms", xlab="Bedrooms", ylab="House Price")
model_HouseBedrooms <- lm(Price ~ Avg..Area.Number.of.Bedrooms, data = housing)
abline(model_HouseBedrooms, col="red")
```

```{r}
plot(housing$Area.Population, housing$Price, main="House Price vs Population", xlab="Population", ylab="House Price")
model_Population <- lm(Price ~ Area.Population, data = housing)
abline(model_Population, col="red")
```

We can see from the graphs that "Average Income" has the most significant increase in price with "Average House Age" being the second most impactful.

There is a slight increase in price with the other 3 variables but it is minimal.

# Conclusion?

Data shows that the more money people earn the more they're willing to spend on housing.

There are outliers but on average a person spends roughly:

```{r}
housing$Percentage <- (housing$Price / housing$Avg..Area.Income) * 100
average_percentage <- mean(housing$Percentage)
average_percentage
```
-Percent of their yearly income on housing.

This means that in order to buy the houses they need to save up for 17.88 years of their salary or get a house loan with a duration of 20 years or longer.



